Data Management and Domain Adjusting Strategies for Implementing Parallel SPH Method in JPUE Simulation

### 2016-11-30 Paper for submission.tex
a new branch

% author names and affiliations
% use a multiple column layout for up to three different
% affiliations
\author{\IEEEauthorblockN{Zhixuan Cao, Abani Patra}
\IEEEauthorblockA{Department of MAE \\%Mechanical and Aerospace Engineering
 The University at Buffalo\\
 Buffalo, New York 14260\\
Email: zhixuanc@buffalo.edu,\\abani@buffalo.edu}
\and
\IEEEauthorblockN{Matthew Jones}
\IEEEauthorblockA{Center for Computational Research\\%Computer Science and Engineering
The University at Buffalo\\
Buffalo, New York 14260\\
Email: jonesm@buffalo.edu}
\and
\IEEEauthorblockN{Jaroslaw Zola}
\IEEEauthorblockA{Department of CSE\\
The University at Buffalo\\
Buffalo, New York 14260\\
Email: jzola@buffalo.edu}
}
% conference papers do not typically use \thanks and this command
% is locked out in conference mode. If really needed, such as for
% the acknowledgment of grants, issue a \IEEEoverridecommandlockouts
% after \documentclass

@@

% As a general rule, do not put math, special symbols or citations
% in the abstract
\begin{abstract}
This paper presents a parallel implementation of smoothed particle hydrodynamics (SPH) method using the message passing interface (MPI) standard to simulate a turbulent jet or plume which is ejected from a nozzle into a uniform environment(JPUE). Background grid is used to reduce neighbors searching cost and to decompose (redecompose) the domain. Space Filing Curve(SFC) based index is adopted to assign an unique identifier to each background grid. As simulation of JUPE requires adding of new particles during simulation, time-dependent SFC based indexes are assigned to particles to guarantee uniqueness of identifiers.  
Both particle and background grid are managed by hash tables which ensure quick and flexible accessing, adding and deleting of element.
A SFC based three dimensional domain decomposition and a dynamic load balancing strategy are implemented to ensure good load balance. In addition, computational domain is adjusted during simulation to reduce computational cost. 
Numerical tests show that our code has acceptable strong scalability and good weak scalability.
These strategies can be further applied to many other implementations of meshfree methods, especially those implementations that require irregular
particle distributions, adding and deleting of particles and leverage adaptive background grids.
\end{abstract}

@@

SPH is a meshless scheme, initially developed for astrophysical applications by Lucy\cite{lucy1977numerical} and Gingold and Monaghan\cite{gingold1977smoothed}. 

@@

Goozee\cite{goozee2003distributed} implemented a simple SPH code using MPI, OpenMP and BSP. 
Wenbo\cite{wenbo2014performance} presented a parallel SPH implementation on multi-core CPUs. 
Holmes\cite{holmes2011framework} presented a simulation framework that enables distributed numerical computing in multi-core shared-memory environments. 
Dominguez\cite{dominguez2011optimization} presented optimizations for both CPU and GPU parallelization of a SPH method. 
Ferrari\cite{ferrari2009new} parallelized a 3D SPH code using the message passing interface (MPI) standard, together with a dynamic load balancing strategy to improve the computational efficiency of the scheme. 

@@

Crespo\cite{crespo2015dualsphysics} used the GPUs to accelerate DualSPHysics by up to two orders of magnitude compared to the performance of the serial version.\\

@@

simple scenarios like breaking-waves, floodings etc. Work on more complicated problems, such as eruption of volcano plume (figure \ref{fig:Plume}) which is essentially a multiphase turbulent flow,

@@

Prediction of such complicated phenomena with acceptable accuracy at given time window requires resolutions (very high particle counts) that cannot been accomplished without parallel computing. What's more, imposing of some types of boundary conditions (such as realistic wind field, eruption boundary condition) requires dynamically adding and removing of particles during simulation. This requires efficient and more flexible data management scheme. What we need to point out is that the benefit of flexible data management strategies are not only limited to specific implementations of SPH. Actually, flexibility in data management are more critical for several advanced techniques of SPH, such as dynamic particle splitting techniques\cite{vacondio2012accurate, feldman2007dynamic}, which will give greater resolution at the area of interest by splitting one large particle to several smaller ones.

@@

Among existing CPU parallel SPH schemes, most of them focus on neighbors searching algorithm and dynamic load balancing. (eg.\cite{ferrari2009new, crespo2015dualsphysics}). Less attention has been paid to developing of more flexible data management schemes for more complicated problems. 
Fortunately, efficient and flexible data management strategies for high performance computing have been successfully implemented in mesh based methods (eg. \cite{laszloffy2000simple} for adaptive hp FEM, and \cite{pitman2003computing, patra2005parallel} for FVM). Motivated by techniques developed for mesh based methods, we present a complete framwork for parallelizing SPH program with MPI standard model allowing more flexible and efficient data access in this paper.\\
Any implementation of SPH code requires efficient searching and updating of neighbors during simulation. Of the many choices possible for this we adopted a background grid which was proposed by Monaghan and Lattanzio\cite{monaghan1985refined} and is quite popular in parallel SPH. The background grid is also used for domain decomposition in SPH. We refer to the elements of background grid, namely squares for two dimension and cubics for three dimension, as buckets. 

@@

In both SPHysics and DualSPHysics\cite{crespo2015dualsphysics}, the physical quantities of each particle (position, velocity, density...) are stored in arrays, and the particles (and the arrays with particle data) are reordered following the order of the cells. This has two advantages: 1) access pattern is more regular and more efficient, 2) it is ease to identify the particles that belong to a cell by using a range since the first particle of each cell is known. But adding, deleting and especially accessing of particles are cumbersome. Ferrari\cite{ferrari2009new} adopted linked lists using pointers so that particles can be deleted or added during the simulation. Storage problems caused by fix-size arrays are thereby also eliminated. We define C++ classes which contain all data of particle and bucket. As for the management of data, we adopt hash tables to store pointers to particles and buckets, which give us not only flexibility of deleting and adding element, but also quicker access compared with linked list. Instead of using the "nature manner" to number particles, we adopt SFC based index to give each particle and background bucket an unique identifier -- a strategy known to preserve some locality at minimal cost. The SFC based numbering strategy is further extended to include time step information so that particles added at the same position but different time will have different identifiers. 

@@

To the best of the author's knowledge, no implementation of SPH has the feature of adjusting computational domain based on simulation needs. For JPUE simulation, such feature will greatly reduce computational cost by avoiding computing of uninfluenced fluid. This feature is accomplished by adding a scan function to monitor the outermost layer of the domain and turn ghost particles to real particles at proper time.\\

@@

The data structure, particle and bucket indexing strategies, domain decomposition, dynamic load balancing method and domain adjusting strategies in this paper can be easily adopted by other implementations of any meshfree methods(include SPH). The flexibility of data accessing enables implementing of meshless methods for solving of more complicated problems and using of more advanced techniques.

@@

and are kept fixed in time during the entire simulation, to reduce searching cost (since search can now be restricted to only neighboring domains).
Domain decomposition will be based on a SFC going through centroids of all buckets. A basic work flow of our SPH code is shown in figure \ref{fig:Work_flow}.\\
Particles need to be added and(or) removed during simulation of JPUE problems. Wall boundary conditions in SPH can be imposed either by adding a force term or using ghost particles. We adopt the latter one in our simulation. As our computational domain will be adjusted during simulation, wall ghost particles need to be added during simulation. New particles also need to be added for the eruption boundary condition. 
We will describe in this section strategies which satisfy these requirements.

@@

We need to emphasize here that in theory motion of particles will destroy locality established based on initial coordinates of particles. However, as particles are moving fairly regularly in a JPUE, the locality of most of particles is largely conserved during simulation. We will revisit this question for more chaotic flows and 

@@

One way to handle hash conflicts is by using an additional sorted vector attached to the hash table. When several keys hash to the same slot, a vector will be created. The vector is sorted based on keys so that a binary search can be used to find the correct position for adding, deleting or retrieving. Another option to handle hash conflicts is using an additional link list which is more flexible in memory allocation. The average time complexity of binary search is O(log n) while that for linear search based on link list is O(n). However, accessing efficiency of link list is much lower than array based data structure, especially when the link list becomes longer. Choosing of proper way to handle hash conflicts greatly depends on the problem itself. For the test problem in this paper, successively adding particles at the bottom of the eject vent will lead to hash conflicts of many particles, which implies a long link list. But considering the very long conflicts only occur on several slots among millions (see figure \ref{fig:Particle_adding_with_link}), we still choose the link list to handle hash conflicts. This decision was made based on numerical experiments.

@@

\subsubsection{Weighted work load}
Particles used in the test problem can be categorized into four types based on the particle-type-flag: real particle, wall ghost particle, pressure ghost particle and eruption ghost particle. Ghost particles are for imposing of corresponding boundary conditions (see figure \ref{fig:Domain_adjusting}). As different types of particles involve different amount of computational work, shown by table \ref{tab:Computational_cost_particles} and table \ref{tab:Computational_cost_steps}, we assign different work load weight for different types of particles based on profilling data. Instead of simply using number of contained particles as work load for bucket, work load of each bucket is determined by summing up work load weight of all particles within the bucket. The SFC sequence passing through centroids of all buckets now becomes a weighted sequence. Domain decomposition  is conducted based on the weighted SFC of buckets.

@@

    Real & Yes & Yes & Yes & Yes & Yes & No & 2.00\\

@@

Figure \ref{fig:SFC_domain_decomposition} shows how domain is decomposed based on partition of SFC of buckets. The particles are automatically split into several groups along with buckets that contain them. As SFC of buckets is a curve in the three dimensional space, partition of this curve will automatically lead to 3D domain decomposition. A 2D domain decomposition based on SFC of footprint buckets projected by three dimensional buckets was adopted by Kumar et al \cite{kumar2013parallel}. A comparison of the scalability of these two schemes (see figure \ref{fig:2D_vs_3D_efficiency}) confirms that 3D domain decomposition is a better choice when processes number is larger. 

@@

%SPH particles can be viewed as discretization points\cite{diltsg1999moving}. 

@@

collective support of the SPH particles defines the computational domain. However, for JPUE simulation (and many similar/related phenomena), where some fluid ejects into stationary fluid and gets mixed, the domain-adjusting feature of SPH cannot be taken advantage of as all of the stationary fluid must be represented. %Because the whole domain, which occupied by stationary fluid before ejected fluid reaching there, has to be discretized at the very beginning of simulation. 
A lot of CPU time will be spent on computing associated with these stationary particles.  If simulating of stationary particles can be avoided, the computational cost will be reduced greatly. %To the best of the author's knowledge, no implementation of SPH has the feature of adjusting computational domain based on simulation. 
We propose a simple strategy to add such feature in our code with low computational cost. We add a scan function to monitor the outermost layer of the domain. When the ejected fluid reaches the boundary of the current domain, ghost particles (for pressure boundary condition) will be turn to real particles and then add new ghost particles for pressure boundary condition. The original work flow (see figure \ref{fig:Work_flow}) is modified to enable such a feature (see figure \ref{fig:Work_flow_adjust}).\\
This is implemented by using a involved-flag, in the particle class. Particles are categorized into three groups based on the value of the involved-flag: involved (involved-flag=2), potentially involved (involved-flag=1), and not involved (involved-flag=0). The involved particles are particles that have already been affected by the mixing. The potentially involved particles are particles that have not been involved in mixing but are adjacent to involved particles and will thus be involved in the near future. %These not involved particles are particles that far away from the eject source and are still at the initial state. For these not involved particles, it is meaningless to update its physical properties. As a consequence, searching for neighbors and data communitation for these particles are also unnecessary.

@@

Our target here is to develop data management and parallelization strategies for more complicated implementations of SPH which demand quick and flexible data access, deletion and addition. Thus, the test problem used should have such requirements. Simulations of a volcanic eruption, which is essentially a multi-phase, turbulent, ejection mixing flow accompanied with microphysics processes will place these demands on the framework. Thus, we adopt a two phase volcanic plume model\cite{suzuki2005numerical} as our test problem. In this model, one phase is air while another phase is ejected material.
\subsection{Governing equations\cite{suzuki2005numerical} and boundary conditions}
Based on Navier-Stokes equations and several simplifications, the governing equations in Eulerian form are:
%the following assumptions are made:
%\begin{itemize}
%\item Molecular Viscosity is neglected since eddy viscosity due to turbulence is dominant. As heat conduction is much smaller than turbulent heat exchange \cite{oberhuber1998volcanic}, We only consider turbulent heat exchange.
%\item Erupted material consist of solid with different size and mixture of gas (various constituent) is assumed to be well mixed and behave like a single phase fluid (phase 2). Air (also a mixture of different gas) is assumed to be another phase (phase 1).
%\item We assume immediate thermodynamic equilibrium so that no separate energy equation is needed for each phase. As a result, there is only one energy equation for both phases and heat transfer effect between different phases will be ignored.
%\item We assume immediate dynamic equilibrium so that no separate momentum equation is needed for each phase. As a result, there is only one vector moment equation for both phases. Drag force term will not show up in the momentum equation as two phases always move with the same velocity.
%\item Because of above assumptions, all other micro-physics process (like phase change of $H_2O$, aggregation, decomposition, absorption of gas on the surface of solid, solution of gas into the liquid) and chemical process will not be considered in this model.
%\end{itemize}
%Finally the governing equations for this model in Eulerian form are:
\begin{eqnarray} %%I did not notice that I can add equation like this, useful, haha
\dfrac{\partial \rho}{\partial t} + \nabla \cdot (\rho \textbf{v}) &=& 0 \label{eq:gov-cs-rho} \\
%\end{equation}
%\begin{equation}
\dfrac{\partial \rho \xi}{\partial t} + \nabla \cdot (\rho \xi \textbf{v}) &=& 0 \label{eq:gov-cs-ks}\\
%\end{equation}
%\begin{equation}
\dfrac{\partial \rho \textbf{v}}{\partial t} + \nabla \cdot (\rho \textbf{v} \textbf{v} + p\textbf{I}) &=& \rho \textbf{g} \label{eq:gov-cs-v} \\
%\end{equation}
%\begin{equation}
\dfrac{\partial \rho E}{\partial t} + \nabla \cdot [(\rho E + p )\textbf{v}] &=& \rho \textbf{g} \cdot\textbf{v} \label{eq:gov-cs-e}
\end{eqnarray}
$\xi$ is the mass fraction of ejected material.
$E = e + K $ is total energy which is summation of kinetic energy $K$ and internal energy $e$.
An additional equation is required to close the system. In this model, the equation for closing the system is an EOS:
\begin{align}
p = (\gamma_m - 1)\rho e \label{eq:EOS}
\end{align}
Where 
\gamma_m = R_m/C_{vm} + 1 \label{eq:gov-gm}
R_m = n_gR_g + n_aR_a  \label{eq:gov-Rm}
C_{vm} = n_s C_{vs} + n_g C_{vg} + n_a C_{va} \label{eq:gov-Cvm}
n_a = 1 - \xi \label{eq:gov-na}
n_g = \xi n_{g0} \label{eq:gov-ng}
n_s = \xi - n_g \label{eq:gov-ns}
Where, $C_v$ is specific heat with constant volume, $n$ is mass fraction, $R$ is gas constant. The subscript 
$m$ represents mixture of ejected material and air, $s$ is solid portion in ejected material, $g$ is gas portion in the ejected material and $a$ is air.\\
In current model the initial domain is a 3D box. 
%The boundaries are categorized into eruption vent (a circle area at the center of the bottom), wall boundary (box bottom), pressure boundary (Other faces of the box).
At the eruption vent, temperature $T$, velocity $\textbf{v}$, pressure $p$, initial mass fraction of gas portition $n_{g0}$ and mass discharge rate $\dot M$ is given. 
%The radius of vent is determined from $\rho$, $\dot M$ and $\textbf{v}$.
Non-slip, adiabatic wall boundary condition is imposed on the wall. The pressure of the surrounding atmosphere is specified, forming pressure boundary conditions. 
%Except for the pressure, density, velocity, and energy will depend on the solution. We adpot ghost particles to impose these different type of boundary conditions (see figure \ref{fig:Domain_adjusting}). 
\subsection{Discretized governing equations with SPH}
There are several review papers \cite{monaghan1992smoothed, monaghan2005smoothed, price2012smoothed, monaghan2012smoothed} which provide a pretty comprehensive descriptions of SPH. These may be consulted for additional details. %We will not cover the basic theory of SPH in this paper.
The discretized governing equations with SPH are:
<\rho_a^a>=\sum m_b w_{ab} (h_a) \label{eq:gov-sph-d1}
<\rho_i^{sg}>=\sum_j m_j w_{ij} (h_i) \label{eq:gov-sph-d2}
<\dfrac{d \textbf{v}_{\alpha}}{d t}>= &
-\sum_b [m_b (\dfrac{p_b}{\rho_b^2} + \dfrac{p_{\alpha}}{\rho_{\alpha}^2} + \Pi_{\alpha b}) \\ &\bigtriangledown_{\alpha}w_{\alpha b}(h_{\alpha})]
-\sum_j [m_j (\dfrac{p_j}{\rho_j^2} \\ & + \dfrac{p_{\alpha}}{\rho_{\alpha}^2} + \Pi_{\alpha j}) \bigtriangledown_{\alpha}w_{\alpha j}(h_{\alpha})]
+\textbf{g}
\label{eq:gov-sph-v}
<\dfrac{d e_{\alpha}}{d t}>=&
 0.5\sum_b [m_b \textbf{v}_{\alpha b}(\dfrac{p_b}{\rho_b^2} + \dfrac{p_{\alpha}}{\rho_{\alpha}^2} + \Pi_{\alpha b})\\ & \bigtriangledown_{\alpha}w_{\alpha b}(h_{\alpha})]
+0.5\sum_j [m_j \textbf{v}_{\alpha b}(\dfrac{p_j}{\rho_j^2} \\ & + \dfrac{p_{\alpha}}{\rho_{\alpha}^2} + \Pi_{\alpha j}) \bigtriangledown_{\alpha}w_{\alpha j}(h_{\alpha})]
\label{eq:gov-sph-e}
Where
$\rho_a^a$ is density of phase 1 (air).
$\rho_i^{sg}$ is density of phase 2 (erupted material).
$\rho=\rho^a + \rho^{sg}$ is density of mixture of phase 1 and phase 2.
\textbf{v}_{\alpha \beta}=\textbf{v}_{\alpha}-\textbf{v}_{\beta}
$w_{ab} (h_a)= w(\textbf{r}_a-\textbf{r}_b,h_a)$ is smoothing kernel.
%$\bigtriangledown_{\alpha}w_{\alpha j}(h_{\alpha})$-->need double check
$\Pi$ is artificial viscosity term \cite{monaghan1992smoothed}.
Index $a$, $b$ is for phase 1.
Index $i$, $j$ is for phase 2.
Index $\alpha$, $\beta$ can be index of either phase 1 or phase 2.
The position of each particle is updated according to the following equation.
\begin{align}
\dfrac{d \textbf{r}_a}{dt} = \textbf{v}_{\alpha} \label{eq:gov-update-pos}
\end{align}
Free surface flows are inherently turbulent. We adopt $SPH-\varepsilon$ method developed by Monaghan\cite{monaghan2011turbulence} to capture turbulence in the plume. This will result in a filtered velocity for position updating and additional turbulent terms in momentum and energy equation.
\subsection{Solver performance}
\includegraphics[scale=0.33]{2cases_time}
\caption{Wall clock time of test case1 and test case2}
\label{fig:2cases_time}
\includegraphics[scale=0.33]{2cases_efficiency}
\caption{Influence of total work load on strong scalability}
\includegraphics[scale=0.33]{2D_vs_3D_efficiency}
\caption{Strong scalability of 3D domain decomposition and 2D domain decomposition}
\label{fig:2D_vs_3D_efficiency}
\includegraphics[scale=0.35]{weak_loglog}
Three types of processors -- 8 core L5520/L5630 running at 2.27/2.13GHz clock rate with 3GB memory per core on a Mellanox Infiniband network and 12 core E5645 running at 2.40GHz clock rate with 4GB memory per core on a Q-Logic Infiniband network. Each node comprise of two sockets with several of these processors. Memory and level 3 cache are shared on each node. 
The initial domain is $[-4.8km,4.8km]\times [-4.8km,4.8km] \times [0km, 6km]$, with smoothing length (we set initial intervals between particles equal to smoothing length) equals to 200m and 100m respectively for test case1 and test case2. The computational work load of test case 2 is 8 times of that of the test case 1. The simulations run for 20s physical time.  Wall clock time and speed up are shown in figure \ref{fig:2cases_time} and \ref{fig:2cases_efficiency}. Linear speed up is observed when number of processes is smaller than 16. Test case2 shows better speed up than test case1 which implies that the overhead of strong scalability can be increased by increasing total amount of work load. We also compared the performance of 2D domain decomposition and 3D domain decomposition. As shown by figure \ref{fig:2D_vs_3D_efficiency}, 2D domain decomposition shows a little bit better speed up than 3D domain decomposition for 32 processes. This can be explained by the fact that 2D domain decomposition can get more regular subdomain and as a results, need less communication. When number of processes increase, it becomes harder for 2D decomposition to balance workload, that is why 3D domain decomposition has better speed up for processes number equals to 64 and 128.\\ 
The weak scalability test is conducted with the same initial domain and various smoothing length. Each simulation runs for 400 loops. The average number of real particles of each process keeps constant at $25900$. 
Tests are carried out under two different settings: the normal scheduling (hence readily available) wherein the nodes may be shared with other users and exclusive access setting. Results of normal setting tests may suffer from contention for memory, CPU-memory bandwidth and network access but are representative of what will be available during production use.
As shown in figure \ref{fig:weak_loglog}, simulation time are almost constant with some minor fluctutations except for an abrupt jump from 132 processors to 620 processors. This jump, based on monitor data from nodes health check system on CCR, is due to a daemon which eats up a cpu. This inteference happens for all tests whose number of processes larger than or equal to 620. The wall clock time of exclusive test is slightly smaller than non-exclusive test which consistent with our anticipation. 
%These fluctuations are due to sharing nodes with other users. In addition, difference in simulation time for different processors are obvious.
%That can be explained by the fact that the total number of footprint buckets in x-y plane is 64 in the test case.
%When we request 128 processes, 64 processes will be totally idle at the beginning. As simulation goes, the domain will expand and more and more idle processes will start getting tasks. That's why we still get some speed up when number of process is 128. 3D domain decomposition is able to divide the work load in a more flexible way and achieve better strong scalability.
For the test problem in this paper, the volcanic plume will finally reach to a region of $[-10km \,\,\, 10km] \times [-10km\,\,\,10km] \times [0km\,\,\,20km]$ after around 300 seconds of eruption. When numerical simulation goes up to 90 seconds, the plume is still within a region of $[-3km\,\,\,3km] \times [-3km\,\,\,3km] \times [0km\,\,\,6km]$. This implies that adjusting of domain can avoid computing large number of uninfluenced air particles, especially for the beginning stage of simulation. Numercial test shows that simulation time of the test problem is reduce to $\frac{1}{4}$ of original simulation time when we adopt the domain adjusting strategy in our code. 
A typical simulation result of volcano plume is shown in figure \ref{fig:Plume}. Input parameters for this simulation are exactly the same as "Run P1" in reference \cite{suzuki2005numerical}.

@@

>>>>

% author names and affiliations
% use a multiple column layout for up to three different
% affiliations
\author{\IEEEauthorblockN{Zhixuan Cao, Abani Patra}
\IEEEauthorblockA{Department of MAE \\%Mechanical and Aerospace Engineering
 The University at Buffalo\\
 Buffalo, New York 14260\\
Email: zhixuanc@buffalo.edu,\\abani@buffalo.edu}
\and
\IEEEauthorblockN{Matthew Jones}
\IEEEauthorblockA{Center for Computational Research\\%Computer Science and Engineering
The University at Buffalo\\
Buffalo, New York 14260\\
Email: jonesm@buffalo.edu}
%\and
%\IEEEauthorblockN{Jaroslaw Zola}
%\IEEEauthorblockA{Department of CSE\\
%The University at Buffalo\\
%Buffalo, New York 14260\\
%Email: jzola@buffalo.edu}
}
% conference papers do not typically use \thanks and this command
% is locked out in conference mode. If really needed, such as for
% the acknowledgment of grants, issue a \IEEEoverridecommandlockouts
% after \documentclass

@@

% As a general rule, do not put math, special symbols or citations
% in the abstract
\begin{abstract}
This paper presents a parallel implementation of smoothed particle hydrodynamics (SPH) method using the message passing interface (MPI) standard to simulate a turbulent jet or plume which is ejected from a nozzle into a uniform environment (JPUE). Background grid is used to reduce neighbors searching cost and to decompose (redecompose) the domain. Space Filing Curve (SFC) based index is adopted to assign an unique identifier to each background grid. As simulation of JPUE requires adding of new particles during simulation, time-dependent SFC based indexes are assigned to particles to guarantee uniqueness of identifiers.  
Both particle and background grid are managed by hash tables which ensure quick and flexible accessing, adding and deleting of elements.
A SFC based three dimensional domain decomposition and a dynamic load balancing strategy are implemented to ensure good load balance. In addition, computational domain is adjusted during simulation to reduce computational cost. 
Numerical tests show that our code has acceptable strong scalability and good weak scalability.
These strategies can be further applied to many other implementations of meshfree methods, especially those implementations that require irregular particle distributions, adding and deleting of particles and leverage adaptive background grids.
\end{abstract}

@@

SPH is a meshless scheme, initially developed for astrophysical applications by Lucy \cite{lucy1977numerical} and Gingold and Monaghan \cite{gingold1977smoothed}. 

@@

Goozee \cite{goozee2003distributed} implemented a simple SPH code using MPI, OpenMP and BSP. 
Wenbo \cite{wenbo2014performance} presented a parallel SPH implementation on multi-core CPUs. 
Holmes \cite{holmes2011framework} presented a simulation framework that enables distributed numerical computing in multi-core shared-memory environments. 
Dominguez \cite{dominguez2011optimization} presented optimizations for both CPU and GPU parallelization of a SPH method. 
Ferrari \cite{ferrari2009new} parallelized a 3D SPH code using the message passing interface (MPI) standard, together with a dynamic load balancing strategy to improve the computational efficiency of the scheme. 

@@

Crespo \cite{crespo2015dualsphysics} used the GPUs to accelerate DualSPHysics by up to two orders of magnitude compared to the performance of the serial version.\\

@@

simple scenarios like breaking-waves, floodings etc. Work on more complicated problems, such as eruption of volcano plume (Fig. \ref{fig:Plume}) which is essentially a multiphase turbulent,

@@

Prediction of such complicated phenomena with acceptable accuracy at given time window requires resolutions (very high particle counts) that cannot been accomplished without parallel computing. What's more, imposing of some types of boundary conditions (such as realistic wind field, eruption boundary condition) requires dynamically adding and removing of particles during simulation. This requires efficient and more flexible data management scheme. What we need to point out is that the benefit of flexible data management strategies are not only limited to specific implementations of SPH. Actually, flexibility in data management are more critical for several advanced techniques of SPH, such as dynamic particle splitting techniques\cite{vacondio2012accurate, feldman2007dynamic}, which will give higher resolution at the area of interest by splitting one large particle to several smaller ones.

@@

Among existing CPU parallel SPH schemes, most of them focus on neighbors searching algorithm and dynamic load balancing. (eg. \cite{ferrari2009new, crespo2015dualsphysics}). Less attention has been paid to developing of more flexible data management schemes for more complicated problems. 
Fortunately, efficient and flexible data management strategies for high performance computing have been successfully implemented in mesh based methods (eg. \cite{laszloffy2000simple} for adaptive hp FEM, and \cite{pitman2003computing, patra2005parallel} for FVM). Motivated by techniques developed for mesh based methods, we present a complete framwork for parallelizing SPH program with MPI standard model allowing more flexible and efficient data access.\\
Any implementation of SPH code requires efficient searching and updating of neighbors during simulation. Of the many possible choices we adopt a background grid which was proposed by Monaghan and Lattanzio\cite{monaghan1985refined} and is quite popular in parallel SPH. The background grid is also used for domain decomposition in SPH. We refer to the elements of background grid, namely squares for two dimension and cubics for three dimension, as buckets. 

@@

In both SPHysics and DualSPHysics\cite{crespo2015dualsphysics}, the physical quantities of each particle (position, velocity, density...) are stored in arrays, and the particles (and the arrays with particle data) are reordered following the order of the cells. This has two advantages: 1) access pattern is more regular and more efficient, 2) it is easy to identify the particles that belong to a cell by using a range since the first particle of each cell is known. But adding, deleting and especially accessing of particles are not flexible enough. Ferrari\cite{ferrari2009new} adopted linked lists using pointers so that particles can be deleted or added during the simulation. Storage problems caused by fix-size arrays are thereby also eliminated. We define C++ classes which contain all data for particle and bucket. As for the management of data, we adopt hash tables to store pointers to objects of particles and buckets, which give us not only flexibility of deleting and adding element, but also quicker access compared with linked list. Instead of using the "nature manner" to number particles, we adopt SFC based index to give each particle and background bucket an unique identifier -- a strategy known to preserve some locality at minimal cost. The SFC based numbering strategy is further extended to include time step information so that particles added at the same position but different time will have different identifiers. 

@@

To the best of the author's knowledge, no implementation of SPH has the feature of adjusting computational domain based on simulation needs. For JPUE simulation, such feature will greatly reduce computational cost by avoiding computing of uninfluenced fluids. This feature is accomplished by adding a scan function to monitor the outermost layer of the domain and turn ghost particles to real particles at proper time.\\

@@

The data structure, particle and bucket indexing strategies, domain decomposition, dynamic load balancing method and domain adjusting strategies in this paper can be easily adopted by other implementations of any meshfree methods (include SPH). The flexibility of data accessing enables implementing of meshless methods for solving of more complicated problems and using of more advanced techniques.

@@

and are kept fixed in time during the entire simulation, to reduce searching cost (since search can now be restricted to only neighboring domains). A basic work flow of our SPH code is shown in figure \ref{fig:Work_flow}.
Domain is decomposed based on a SFC going through centroids of all buckets. Our data management strategies will also guarantee flexible adding, deleting and accessing of data.
%Wall boundary conditions in SPH can be imposed either by adding a force term or using ghost particles. We adopt the latter one in our simulation. As our computational domain will be adjusted during simulation, wall ghost particles need to be added during simulation. New particles also need to be added for the eruption boundary condition. 
%We will describe in this section strategies which satisfy these requirements.

@@

We need to emphasize here that, in theory, motion of particles will destroy locality established based on initial coordinates of particles. However, as particles are moving fairly regularly in a JPUE, the locality of most of particles is largely conserved during simulation. We will revisit this question for more chaotic flows and 

@@

One way to handle hash conflicts is using an additional sorted vector attached to the hash table. When several keys hash to the same slot, a vector will be created. The vector is sorted based on keys so that a binary search can be used to find the correct position for adding, deleting or retrieving. Another option to handle hash conflicts is using an additional link list which is more flexible in memory allocation. The average time complexity of binary search is O(log n) while that for linear search based on link list is O(n). However, accessing efficiency of link list is much lower than array based data structure, especially when the link list becomes longer. Choosing of proper way to handle hash conflicts greatly depends on the problem itself. For the test problem in this paper, successively adding particles at the bottom of the eject vent will lead to hash conflicts of many particles, which will lead to a long link list. But considering the conflicts only occur on several slots among millions (see figure \ref{fig:Particle_adding_with_link}), the link list is a better choice for us. 
%This decision was made based on numerical experiments.

@@

\subsubsection{Weighted work load} \label{sec:Weighted_work_load}
Particles used in the test problem can be categorized into four types based on the particle-type-flag: real particle, wall ghost particle, pressure ghost particle and eruption ghost particle. Ghost particles are for imposing of corresponding boundary conditions (see figure \ref{fig:Domain_adjusting}). As different types of particles involve different amount of computational work, shown by table \ref{tab:Computational_cost_particles} and table \ref{tab:Computational_cost_steps}, we assign different work load weight for different types of particles calibrated by profilling data. Instead of simply using number of contained particles as work load for bucket, work load of each bucket is determined by summing up work load weights of all particles within the bucket. The SFC sequence passing through centroids of all buckets now becomes a weighted sequence. Domain decomposition  is conducted based on the weighted SFC of buckets.

@@

    Real & Yes & Yes & Yes & Yes & Yes & No & 1.98\\

@@

Figure \ref{fig:SFC_domain_decomposition} shows how domain is decomposed based on partition of SFC of buckets. The particles are automatically split into several groups along with buckets that contain them. As SFC of buckets is a curve in the three dimensional space, partition of this curve will automatically lead to 3D domain decomposition. 
%A 2D domain decomposition based on SFC of footprint buckets projected by three dimensional buckets was adopted by Kumar et al \cite{kumar2013parallel}. A comparison of the scalability of these two schemes (see figure \ref{fig:2D_vs_3D_efficiency}) confirms that 3D domain decomposition is a better choice when processes number is larger. 

@@

@@

collective support of the SPH particles defines the computational domain. However, for JPUE simulation (and many similar/related phenomena), where some fluid ejects into stationary fluid and gets mixed, the domain-adjusting feature of SPH cannot be taken advantage of as all of the stationary fluid must be represented.
A lot of CPU time will be spent on computing associated with these stationary particles.  If simulating of stationary particles can be avoided, the computational cost will be reduced greatly.
We propose a simple strategy to add such feature in our code with low computational cost. We add a scan function to monitor the outermost layer of the domain. When the ejected fluid reaches the boundary of the current domain, ghost particles (for pressure boundary condition) will be turn to real particles and then add new ghost particles for boundary conditions. The original work flow (see figure \ref{fig:Work_flow}) is modified to enable such a feature (see figure \ref{fig:Work_flow_adjust}).\\
This is implemented by using a involved-flag, in the particle class. Particles are categorized into three groups based on the value of the involved-flag: involved (involved-flag=2), potentially involved (involved-flag=1), and not involved (involved-flag=0). The involved particles are particles that have already been affected by the mixing. The potentially involved particles are particles that have not been involved in mixing but are adjacent to involved particles and will thus be involved in the near future. 

@@

@@

Our target here is to develop data management and parallelization strategies for more complicated implementations of SPH which demand quick and flexible data access, deletion and addition. Thus, the test problem used should have such requirements. Simulations of a volcanic eruption, will place these demands on the framework. Thus, we adopt a two phase volcanic plume model\cite{suzuki2005numerical} as our test problem. Governing equation, boundary condition and input parameters can be found in \cite{suzuki2005numerical}. Details about discretization of the governing equations will not be covered in this paper, neither. A typical simulation result of volcano plume is shown in figure \ref{fig:Plume}. Input parameters for this simulation are exactly the same as "Run P1" in reference \cite{suzuki2005numerical}.
\subsection{Scalability test}
\includegraphics[scale=0.33]{strong_scale}
\caption{Influence of total work load on strong scalability. It will exceed the time limit on CCR to do a serial simulation for case 2, so the serial simulation time is estimated by assuming that speed up of case 1 and case 2 are the same at 64 processors.}
\includegraphics[scale=0.35]{weak_scale}
Three types of processors -- 8 core L5520/L5630 running at 2.27/2.13GHz clock rate with 3GB memory per core on a Mellanox Infiniband network and 12 core E5645 running at 2.40GHz clock rate with 4GB memory per core on a Q-Logic Infiniband network. Each node comprise of two sockets with several of these processors. Memory and level 3 cache are shared on each node. The simulational domain is a box with initial ddimension $[-4.8km, 4.8km]\times [-4.8km, 4.8km] \times [0km, 6km]$ for case 1 and $[-10km, 10km]\times [-10km, 10km] \times [0km, 5km]$ for case 2. The smoothing length (we set initial intervals between particles equal to smoothing length) equals to 200m and 100m respectively for test case 1 and test case 2. So the computational work load of test case 2 is larger than that of the test case 1. The simulations run for 20s physical time.  Parallel speed up are shown in Fig. \ref{fig:2cases_efficiency}. Test case2 shows better speed up than test case1 which implies that the overhead of strong scalability can be increased by increasing total amount of work load.
The weak scalability test is conducted with the same initial domain as test case 1 and various smoothing length. Each simulation runs for 400 time steps. The average number of real particles of each process keeps constant at $25900$, so the average work load for each processor are keep constant. 
As shown in figure \ref{fig:weak_loglog}, simulation time are almost constant with some minor fluctutations.\\
%\includegraphics[scale=0.33]{2D_vs_3D_efficiency}
%\caption{Strong scalability of 3D domain decomposition and 2D domain decomposition}
%\label{fig:2D_vs_3D_efficiency}
